Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-08-13 analysis/paper/paper.Rmd
2nd edit

---
title: "Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less pronounced than island Southeast Asia and Oceania. Throughout East Asia direct colonial rule was rare and limited, but the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused significant transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. The agency of indigenous peoples in the colonial contexts has been broadly discussed to understand the negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Rubertone2000; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005]. Studying the consumption of foreign goods is one way to understand indigenous experiences of colonial encounters and explore the agency of indigenous peoples by examining how goods were distributed and used [@Dietler2005; @Mullins2011; @Scaramelli2005].

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 study both archaeological and historical records of Heping Dao in northern Taiwan that was the location of the colony of San Salvador founded by the Spanish in 1626 and took over by the Dutch in 1642. They uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. It shows a direct impact on the local indigenous populations by disturbing their long settlement pattern and changing local landscape. Different from the view of a Spanish fort, Yilan in Northeastern Taiwan is an especially unique location for exploring indirect colonial effects on local indigenous settlements due to its physically isolated location. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. The sea provides a way through which indigenous people in Yilan could interact with other groups from different regions, including other native indigenous groups, foreign traders such as the Han people of China, and Europeans. Large-scale and frequent trade activities brought a wide variety of trade objects into indigenous communities in Yilan through local regional exchange network. One of the most common types of trade goods in this region were personal ornaments such as glass and stone beads was one of the common foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments are a social signal of an individual’s status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore the evidence from Yilan spanning from late Iron Age to historical period to address the question of whether indirect colonial influences on indigenous populations can be detected. I focus on the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Prior to the contact with the Europeans in the early 17th century, Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems since tenth century, and encountered large waves of Han Chinese immigration after 17th century since Taiwan was incorporated into the territories of Qing Dynasty of China [@Liu2017; @Andrade2007, ch. 11]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf.@Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine their potential of the ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new spatial patterns in the distributions of ornaments at Kiwulan that indicates increasing social inequality. I hypothesize that after Chinese contact there was a decline in the use and discard of ornaments at Kiwulan due to overall population declines and acculturation in the Chinese community.

@@

Northeastern Taiwan was part of the local exchange network in northern Taiwan, where the indigenous people had already established their own exchange network before European arrived that has been described as “inter-insular trade” by Chen @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis. The exchange network in northern Taiwan was increasingly influenced by long-distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since the 15th century due to its location on the shipping routes [@Chen2005]. This was intensified in the 17th century with the arrival of the Europeans and their trade goods [@Wang2007]. Due to its location on the periphery of major trade ports, Keelung and Tamsui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for a religious purpose [@Blusse2000, pp. 343]. The report of the Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. Esquivel also recorded how indigenous people viewed agate pieces as prestige goods in their culture. An indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing[@Borao2009, pp. 122-151]. An indigenous funeral process was recorded, documenting the use of agate beads in ritual contexts, with more agate beads, pottery, and cloths placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. Those historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

@@

Chinese historical records from the year of 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan had the custom to wear ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, a field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called “subli”, by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. It was usually the female shamans conducting divination rituals that passed the beads down from their ancestors. Despite most Kavalan people today not knowing the specific origin of their agate beads, the value of beads is still highlighted through oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to the European contact period, there are fewer mentions of beads in Chinese contact period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.  
# Excavations at Kiwulan in northeastern Taiwan

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text", out.width= "400px", echo=FALSE}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Each square is 4 x 4 m", out.width= "70%", echo=FALSE}

# add to the map and scale and north arrow. 
knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.png"))
```

Kiwulan (Figure \@ref(fig:KWL-map)) is located at northern Yilan and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with a hundred year intervals according to original radiocarbon dates, excavation depth, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5 and L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. I take approximate indicators of the start of European contact at Kiwulan as the appearance of An-ping jars and stonewares (martavans/martaban) that were largely introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages, and the jar shapes found at Kiwulan are typical of those found elsewhere in V.O.C sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods. European colonization ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. I recognize this as the Chinese contact period, which is archaeologically visible at Kiwulan by a large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration can be seen from the Chinese official records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan from sampling area", out.width='60%'}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

# making table
knitr::kable(ornaments_shape,
             caption = "A table of ornament subtypes")

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole area, 3173 pieces from burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments because mostly beads were preserved well in the form of bead strings that consisted of hundreds to thousand pieces. In this paper I focus on the post hole area where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole area represents Kiwulan domestic space that provides clear contexts and basis for comparison to see if there is any spatial pattern of distribution. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes. The image for each subtype in each major class is shown in Figure \@ref(fig:ornaments-image).

Ornaments were found in 30 burials from the sampling area as grave goods. Their positions in burials in association with particular parts of human remains indicate the purpose of body adornment. For example, agate beads, golden beads, and small metal rings were often found near skulls that indicate the use as necklaces or earrings, while big metal rings were found wearing on radius and ulna bones as bangles. This pattern corresponds to the texts regarding the body decorations of indigenous culture from Chinese historical documents [@Chen1963; @Ke1993]. Agate beads were mostly found in female burials with an average of 4 to 5 pieces, while golden beads were found in both genders with an average of 2 to 3 pieces.

@@

# Results of hypothesis testing

In this section I report the tests of the three hypotheses about indirect impacts of European colonization on the indigenous community at Kiwulan. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European contact, and a lesser diversity after Chinese contact, I compare frequencies of ornament types between different time periods. To test the prediction that after European contact were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

## Frequency of ornaments between time periods

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequency for the major class of ornaments between different time periods at Kiwulan. The result of chi-square test for the number of ornaments (X-squared = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = `r chisq_test_result_p_value`) shows that there is a significant difference between time period groups. It indicates a general pattern that most ornament types were present before European contact.The frequencies reached a peak during the European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate bead, metal ring, and bell. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then their frequency decreased in the European contact period and was even much less in the Chinese contact period.

@@

The distribution of frequency for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `spearman_result_rho`, p-value = `r spearman_result_p_value`). This indicates that the increases in diversity can be explained by the effects of culture contact instead of natural effects of sample size. According to the frequency, agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. Agate beads can be divided into seven subtypes, in which hexagonal shape is the most common subtype that appeared before European contact and increased significantly during the European contact and then declined in the Chinese contact. The small oval type shows a similar frequency before and during European contact but is not found in the Chinese contact period. Waxy oval beads were only found during European contact period. A higher variety of agate beads were found in European contact compared to other time periods. Similarly, metal rings can be classified into eight subtypes, in which wide small ring is the most common metal ornaments, followed by wide large and thin large shape. The greater varieties for those two classes might indicate multiple origins due to global trade networks stimulated by the Europeans. In contrast, copper bell and glass bead have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of all class of ornament by time periods"}

library(sf)
# read in the spatial data
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

## Spatial distribution of ornament types

>>>>

---
title: "Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

The direct impacts of European colonialism on indigenous communities in East Asia were much less pronounced than island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on indigenous peoples in the periphery of colonial control [@TorrenceandClarke2000; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local indigenous societies caused substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler1997; @Junker1993; @Mitchell2000; @Silliman2005]. Studying the consumption of foreign goods can improve our understanding of the negotiations between colonized and colonizer, and the resistance and accommodations of indigenous people through their daily cultural practices [@Dietler2005; @Dietler2015; @Given2004; @Mullins2011; @Rubertone2000; @Scaramelli2005; @Silliman2001; @Torrence2000; @TorrenceandClarke2000; @Voss2005].

Prior to contact with the Europeans in the early 17th century, Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems since the tenth century, and encountered large waves of Han Chinese immigration after 17th century when Taiwan was incorporated into the territories of Qing Dynasty of China [@Liu2017; @Andrade2007, ch. 11]. The exchange network in northern Taiwan was influenced by long-distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since the 15th century due to its location on shipping routes [@Chen2005]. This intensified in the 17th century with the arrival of the Europeans and their trade goods [@Wang2007].

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is physically isolated and there are no traces of European settlement. This makes Yilan an especially unique location for exploring indirect colonial effects on local indigenous settlements because of isolation. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

\Although mountains isolated Yilan by land, the sea provided a way through which indigenous people in Yilan could interact with groups elsewhere, such as indigenous groups, distant traders such as the Han people of China, and Europeans. Large-scale and frequent trading activities brought a wide variety of objects into indigenous communities in Yilan. \The indigenous people of Yilan established their own exchange network that has been described as “inter-insular trade” by @Chen2005 [p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

In this paper, I explore archaeological evidence from Yilan spanning from the late Iron Age to the historical period to address the question of whether indirect colonial influences on the indigenous populations can be detected. I focus on the ornaments recovered from the archaeological excavations at Kiwulan (1400-1900 AD), a large Iron Age settlement in northeastern Taiwan [@Chen2007]. Northeastern Taiwan provides a good example to explore the indirect colonial impacts on indigenous societies by comparing archaeological evidence from different episodes of culture contact [cf. @Lape2003]. I will describe the variety and cultural context of trade ornaments in this region, and examine the potential of ornaments to inform on social changes relating to trade with colonial groups. I hypothesize that there was a greater diversity of ornaments types and materials at Kiwulan after European contact due to participation in a larger scale exchange network. I also predict that after European contact there will be new patterns in the spatial distributions of ornaments at Kiwulan that indicate increasing social inequality. I hypothesize that after 1662 when European contact declined and Chinese contact became the dominant international tradition partner, there was a decline in the use and discard of ornaments at Kiwulan due to overall population declines and acculturation in the Chinese community. To test the hypothesis of a greater diversity of ornaments types and materials at Kiwulan after European contact, and a lesser diversity after Chinese contact, I compare frequencies of ornament types between different time periods. To test the prediction that after European contact were new spatial patterns in the distributions of ornaments at Kiwulan, I use kernel density and point pattern analyses.

@@

Historical documents from the Spanish and Dutch in the early 17th century provide some information about the social life and culture of indigenous people in Northern Taiwan when the Europeans arrived. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by Dominican priest Fr. Jacinto Esquivel in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, usually used cuentas (agate beads) in exchange for necessities with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use agate as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. Esquivel also recorded how indigenous people viewed agate pieces as prestige goods in their culture. An indigenous man who wanted to get married had to pay agate beads to the parents of his future wife. Also, agate or golden beads could be used to resolve conflicts in their daily lives [@LiandWu2006, pp. 151]. Other records mention that the female shamans (majuorbol) in the tribe would use agate beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of agate beads in ritual contexts, with more agate beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige  [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads or golden beads were already treated as prestige goods before the arrival of the Europeans. These were likely introduced earlier from Chinese traders.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had high value due to its delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, a field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

Recent ethnographic research with the living members of the Kavalan tribe describes how agate beads were used in divination practices, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate beads as valuable objects that are usually passed down from mother to daughter as heirlooms in shaman families, without knowing the exact origin. When not in use, beads were often hidden in safe places, such as on the roof beams. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination, consistent with observations from the European colonial period. Despite most Kavalan people today not knowing the specific origin of their agate beads, the high value of beads is still indicated in oral history and by their scarcity. In sum, we find multiple sources describing how local Indigenous people used ornaments in cultural contexts that represents some social roles or high status. However, compared to the European contact period, there are fewer mentions of beads in Chinese contact period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers.

# Excavations at Kiwulan in northeastern Taiwan

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text", out.width= "400px", echo=FALSE}
# need to adjust Keelung, heping dao, check it out after knitting
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Each square is 4 x 4 m", out.width= "70%", echo=FALSE}
# add to the map and scale and north arrow. Check it out after knitting 
knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.png"))
```

Kiwulan (Figure \@ref(fig:KWL-map)) is located at northern Yilan and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5 and L6), European contact period (L4), and Chinese contact period (L2 and L1).

I identify the start of European contact at Kiwulan at 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. In the archaeological record, the approximate indicators of the start of European contact at Kiwulan are the appearance of An-ping jars and stonewares (martavans/martaban) that were largely introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in V.O.C sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods.

European colonization ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese contact period at Kiwulan is the large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration to Yilan can be seen from official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan", out.width='60%'}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype frequencies at Kiwulan")

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole areas, 3173 pieces from 30 burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments in the form of bead strings that consisted of hundreds to thousands of pieces. In this paper I focus on the post hole areas where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organization. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

@@

# Results of hypothesis testing ?

## Changes in the frequencies of ornament types over time

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequency for the major class of ornaments between different time periods at Kiwulan. The result of chi-square test for the number of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = `r chisq_test_result_p_value`) shows that there is a non-random difference between the three time periods. It indicates that most ornament types were present before European contact. Ornament frequencies reached a peak during European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European contact period and a further decrease in the Chinese contact period.

@@

The distribution of frequency for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p-value = `r spearman_result_p_value`). This indicates that the increases in diversity can be explained by the effects of culture contact instead of effects of sample size. According to the frequency, agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the Europeans. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of all class of ornament by time periods"}
library(sf)
# read in the spatial data
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

## Changes in patterns of the spatial distribution of ornament types

Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all ornaments from the research area for each time period. Before European contact, a greater amount of ornaments were found at the northern and middle parts of the research area. During the European contact period, ornaments were more widespread, with some clusters on the northern part. During the Chinese contact period the distribution is more even again. Figure \@ref(fig:plot-spatial-pattern-each) presents the distribution for the major ornament classes individually, some clusters across the area can be observed during the European contact, such as golden beads and agate beads. However, there seems to be no clear consistent pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, the cluster of golden beads was found at the northern part, while the cluster of agate beads was found in the middle part. In contrast, there are multiple clusters of metal rings that are distributed separately across the research area. Copper bells were usually found individually and seem randomly distributed across the area. In the Chinese Contact period, both the amount and density of different classes of ornaments decreased.
